---
title: "EMAP Streams Project"
---

## Data manipulation libraries and data loading

```{r, message = FALSE, warning = FALSE}
library(readr)
library(dplyr)
library(lubridate)

streams <- read_csv("C:/Users/Konrad/Downloads/data1.csv", na = "NA")
streams <- streams[, -1]
```

## Summarizing the data

```{r}
dim(streams)
#str(streams)
head(streams, 5)
```

## Re-structuring the data slightly

```{r}
#convert dates to numerical dates and adds a column with month data

streams <- streams %>%
  mutate(sample_date = dmy(streams$DATE.x)) %>%
  mutate(sample_month = month(sample_date))

summary(streams$sample_date)
summary(streams$sample_month)
```


## Subsetting the data necessary for ordination
```{r}
#Edited to add afterwards: filtering out NA dates data since we can't use it to divide by season
streams_ordination <- streams %>% 
  filter(!is.na(sample_month)) %>%
  select(site.id:Staphylinidae, sample_date, sample_month)

dim(streams_ordination)
```

## Subset only the insect data for the MDS

```{r}
bugs <- streams_ordination %>% 
  select(Acentrella_insignificans:Staphylinidae)

dim(bugs)
```

## Run the MDS

```{r, message = FALSE, warning = FALSE}
library(vegan)

all_bugs_mds <- metaMDS(bugs, k = 2, trace = T, autotransform = F)
```


## Divide the points by season and then plot them

Dr. Edwards selected rows from the original ordination data and not the "pre-" and "post-" seasonal data. Since there are 12 streams with NAs, I removed these earlier. This way, there are the right number of summer and fall points (391 and 93) in the MDS plot and they're properly filled in. I changed the plotting characters slightly from his code to make it easier to see as well.   
```{r}
summer_points <- all_bugs_mds$points[streams_ordination$sample_month <= 8, ]   
fall_points <- all_bugs_mds$points[streams_ordination$sample_month > 8, ]

plot(all_bugs_mds, type = "n", ylim = c(-1.4, 1.4), xlim = c(-1.4, 1.4), main = "Summer and Fall")
points(summer_points, pch = 1,cex = 2)
points(fall_points, pch = 16,cex = 2)
legend("bottomleft", bty = "n" , pch = c(1,16), c("Summer","Fall"), cex = 1.5 )

```


## Filtering the data by season (not necessary for the MDS, but probably for the ANOSIM): 

```{r}
#391 observations - most of the dataset
summer <- streams_ordination %>%
 filter(sample_month <= 8)

#only generates #93 observations -- leaves 12 NAs 
fall <- streams_ordination %>%
 filter(sample_month > 8)

dim(summer)
dim(fall)

head(summer, 5)
head(fall, 5)

summary(summer$sample_month)
summary(fall$sample_month)

```



```{r}
#didn't get this far, think we would need to create another variable that would replace habitat here
#bugs.anism<-anosim(dist, dta$Habitat)
#bugs.anism

```


