---
title: "R Notebook"
output:
  html_document: 
    fig_height: 6
    fig_width: 9
  html_notebook: default
  pdf_document: default
---

# Data Reading and Input

## Loading libraries
```{r}
library(dplyr)
library(lubridate)
library(vegan)
library(ggplot2)
library(ggrepel)

setwd("C:/Users/Konrad/Desktop/ESM 567/Group Project/Streams Project/Streams Project")

```

## Loading Data

```{r}
varnames <- readr::read_csv("C:/Users/Konrad/Downloads/joined_vectors.csv", col_names = FALSE)

streams <- readr::read_csv("Oregon EMAP Streams Data.csv")

functionalgroups <- readr::read_csv("C:/Users/Konrad/Desktop/ESM 567/Group Project/All Sp Info.csv")
```

## Managing Streams Data

### Dropping unnecessary/duplicated variables 
```{r}
streams <- streams[, -1] #siteid duplicated

streams <- streams %>% #removes 12 sites with NAs
  filter(!is.na(Eco3.x))

streams$LONG.y <- NULL ## Note--Konrad removed this section
streams$Strm_Power.y <- NULL
streams$LAT.y <- NULL
streams$Land_Owner <- NULL

streams$ELEV_m.y <- NULL ## Anna removed these
streams$Elev_FT <- NULL
streams$Area..SQFT. <- NULL
streams$STATION_KEY.1 <- NULL
streams$M_Slp_sqt <- NULL
streams$DATE.y <- NULL
streams$YEAR_EST <- NULL
streams$El_m_sqt <- NULL
streams$LSUB_DMM <- NULL
streams$LSUBD_SD <- NULL

streams$STATION_KEY.x <- NULL
streams$Map_Slope.x <- NULL
streams$X.transitional <- NULL
streams$sample_day <- NULL
streams$Precip_in <- NULL
streams$X.past_hay <- NULL
streams$V1W_MSQ <- NULL
streams$V4W_MSQ <- NULL
streams$Watershed_HUC <- NULL

streams$YEAR <- NULL
streams$REACHLEN <- NULL
streams$V1TM100 <- NULL
streams$sample_month <- NULL
streams$hifrag <- NULL
streams$medfrag <- NULL
streams$lowfrag <- NULL
```


## Changing some variable names so they're more understandable
```{r}
for (i in 1:length(colnames(streams)))
{
  if (length(which(varnames$X1 %in% colnames(streams)[i])) != 0)
  {
      colnames(streams)[i] <- varnames$X2[which(varnames$X1 %in% colnames(streams)[i])]
  }
}
```

## Subdividing into just coastal data

```{r}
coast_data <- streams %>%
  filter(Eco3.x == 1) 

#Adds sample date, sample day, and sample month. Sample day/sample month used for season delineation
coast_data <- coast_data %>%
  mutate(sample_date = dmy(coast_data$DATE.x)) %>%
  mutate(sample_month = month(sample_date)) %>%
  mutate(sample_day = day(sample_date)) %>%
  mutate(Season = case_when(sample_month < 7 | (sample_month == 7 & sample_day <= 20) ~ "Early",
                            (sample_month == 7 & sample_day > 20) | (sample_month == 8 & sample_day <= 20) ~ "Mid",
                            sample_month != "Early" & sample_month != "Mid" ~ "Late")) 
```

## A function to clean the environmental data that we have. Takes the environmental data, drops numeric variables missing more than two sites, and imputes the mean for variables with 1 or 2 missing observations. Drops character -> factor variables that are missing any observations. This function can return only numeric variables, or categorical and numeric variables. This function is called by the subset-data-function which divides data into species or environmental and subsequently needs missing environmental data imputed 

```{r}
clean_data_fn <- function(env_data, numerics = FALSE) {
  numeric_vars <- env_data %>% #select numeric variables
    select_if(is.numeric)
  
  missing_numerics <- which(colSums(is.na(numeric_vars)) <= 2) #keeps numeric variables with <= 2 NAs
  
  no_missing_numerics <- numeric_vars[, missing_numerics]
  
  mean_of_cols <- colMeans(no_missing_numerics, na.rm = TRUE)
  
  for (i in 1:ncol(no_missing_numerics)){ #goes through the variables and adds in the mean if there is an NA
    temp <- which(is.na(no_missing_numerics[, i]))
    if (length(temp) != 0)
    {
      no_missing_numerics[temp, i] <- mean_of_cols[i]
    }
  }
  
  if (numerics == TRUE) #if we're looking just for numeric variables
  {
    return(no_missing_numerics)
  }
  
  char_vars <- env_data %>% #select character variables
    select_if(is.character)
  
  factor_vars <- data.frame(lapply(char_vars, as.factor)) #make them factors
  
  no_missing_factors <- which(colSums(is.na(factor_vars)) == 0) #drop the ones without missing observations
  
  no_missing_vars <- cbind(no_missing_numerics, factor_vars[, no_missing_factors])
  
  return(no_missing_vars) #return the numeric and factor variables
}
```


## This function subsets the streams by season, specifying the season, whether we want species or environmental variables, and whether we just want numeric environmental variables

```{r}
subset_data_fn <- function(data, season, var, numerics = FALSE) {
  season_to_filter <- season
  
  data_name <- data %>%
    filter(Season == season_to_filter)
  
  if (var == "sp") { #select the species we have
    data_name <- data_name %>%
      select(Acentrella_insignificans:Staphylinidae)
    
  } else
  {
    data_name <- data_name %>% #select the environmental/associated variables
      select(site.id, SITE_ID:ncol(coast_data))
    
    if (numerics == TRUE) #check if we want only numeric variables (numerics = TRUE) or numeric and factor (FALSE)
    {
      data_name <- clean_data_fn(data_name, numerics = TRUE)
    } else
    {
      data_name <- clean_data_fn(data_name, numerics = FALSE)
    }
  }
  
  return(data_name)
  
}
```

## Subset the data by season, including into species and environmental data

```{r}
#Species
early_sp <- subset_data_fn(data = coast_data, season = "Early", var = "sp")

#Environmental (categorical + numerical)
early_env <- subset_data_fn(data = coast_data, season = "Early", var = "env", numerics = FALSE)

#Environmental (numerical)
#early_env_numerics <- subset_data_fn(data = coast_data, season = "Early", var = "env", numerics =TRUE)

#Keep only variables that have distinct values (remove variables where each value is identical)
early_env <- early_env[sapply(early_env, function(x) length(unique(x))>1)]
#Remove variables that have duplicated values elsewhere in the dataset
early_env <- early_env[!duplicated(lapply(early_env, summary))]

#early_env_numerics <- early_env_numerics[sapply(early_env_numerics, function(x) length(unique(x))>1)]
#early_env_numerics <- early_env_numerics[!duplicated(lapply(early_env_numerics, summary))]

#Mid
mid_sp <- subset_data_fn(data = coast_data, season = "Mid", var = "sp")
mid_env <- subset_data_fn(data = coast_data, season = "Mid", var = "env", numerics = FALSE)
#mid_env_numerics <- subset_data_fn(data = coast_data, season = "Mid", var = "env", numerics = TRUE)

mid_env <- mid_env[sapply(mid_env, function(x) length(unique(x))>1)]
mid_env <- mid_env[!duplicated(lapply(mid_env, summary))]

#mid_env_numerics <- mid_env_numerics[sapply(mid_env_numerics, function(x) length(unique(x))>1)]
#mid_env_numerics <- mid_env_numerics[!duplicated(lapply(mid_env_numerics, summary))]

#Late
late_sp <- subset_data_fn(data = coast_data, season = "Late", var = "sp")
late_env <- subset_data_fn(data = coast_data, season = "Late", var = "env", numerics = FALSE)
#late_env_numerics <- subset_data_fn(data = coast_data, season = "Late", var = "env", numerics = TRUE)


late_env <- late_env[sapply(late_env, function(x) length(unique(x))>1)]
late_env <- late_env[!duplicated(lapply(late_env, summary))]

#late_env_numerics <- late_env_numerics [sapply(late_env_numerics , function(x) length(unique(x))>1)]
#late_env_numerics <- late_env_numerics[!duplicated(lapply(late_env_numerics, summary))]
```

## Functional Feeding Group Data

```{r}
functionalgroups$Taxon <- toupper(functionalgroups$Taxon)
species_names <- colnames(coast_data[2:244])

copy_coast_data <- coast_data[, 2:244]

colnames(copy_coast_data) <- gsub("_", " ", colnames(copy_coast_data))
colnames(copy_coast_data) <- gsub("[.]", " ", colnames(copy_coast_data))
colnames(copy_coast_data) <- trimws(colnames(copy_coast_data), "both")
colnames(copy_coast_data) <- toupper(colnames(copy_coast_data))
colnames(copy_coast_data) <- gsub(pattern = "\\sGR$", replacement = " GR.", x = colnames(copy_coast_data))


taxa <- gsub("_", " ", species_names)
taxa <- gsub("[.]", " ", taxa)
taxa <- trimws(taxa, "both")
taxa <- toupper(taxa)
taxa <- gsub(pattern = "\\sGR$", replacement = " GR.", x = taxa)
taxa <- data.frame(taxa, "Bugs")
colnames(taxa) = c("Taxon", "Bugs")

matches <- left_join(functionalgroups, taxa, by = c("Taxon" = "Taxon"))

current_matches <- matches %>%
  filter(!is.na(Bugs))

scrapers <- current_matches %>%
  filter(FFG == "SC")

filterers <- current_matches %>%
  filter(FFG == "CF")

predators <- current_matches %>%
  filter(FFG == "PR")

collectors <- current_matches %>%
  filter(FFG == "CG")

omnivores <- current_matches %>%
  filter(FFG == "OM")

shredders <- current_matches %>%
  filter(FFG == "SH")

unknown <- current_matches %>%
  filter(FFG == "UN")

na_ffg <- current_matches %>%
  filter(is.na(FFG))

ph <- current_matches %>% #ph piercer/herbivore
  filter(FFG == "PH")

pa <- current_matches %>% #parasite
  filter(FFG == "PA")

mh <- current_matches %>% #macrophyte herbivore
  filter(FFG == "MH")

season_separator <- function(filter_season)
{
  coast_sp_filt  <- copy_coast_data[, which(colnames(copy_coast_data) %in% filterers$Taxon)]
  coast_sp_pred <- copy_coast_data[, which(colnames(copy_coast_data) %in% predators$Taxon)]
  coast_sp_omni <- copy_coast_data[, which(colnames(copy_coast_data) %in% omnivores$Taxon)]
  coast_sp_shred <- copy_coast_data[, which(colnames(copy_coast_data) %in% shredders$Taxon)]
  coast_sp_scrap <- copy_coast_data[, which(colnames(copy_coast_data) %in% scrapers$Taxon)]
  coast_sp_coll <- copy_coast_data[, which(colnames(copy_coast_data) %in% collectors$Taxon)]
  coast_sp_unkn <- copy_coast_data[, which(colnames(copy_coast_data) %in% unknown$Taxon)]
  coast_sp_na <-  copy_coast_data[, which(colnames(copy_coast_data) %in% na_ffg$Taxon)]
  coast_sp_ph <- copy_coast_data[, which(colnames(copy_coast_data) %in% ph$Taxon)]
  coast_sp_pa <- copy_coast_data[, which(colnames(copy_coast_data) %in% pa$Taxon)]
  coast_sp_mh <- copy_coast_data[, which(colnames(copy_coast_data) %in% mh$Taxon)]
  
  
  abund_filt <- rowSums(coast_sp_filt)
  abund_pred <- rowSums(coast_sp_pred)
  abund_omni <- rowSums(coast_sp_omni)
  abund_shred <- rowSums(coast_sp_shred)
  abund_scrap <- rowSums(coast_sp_scrap)
  abund_coll <- rowSums(coast_sp_coll)
  abund_un <- rowSums(coast_sp_unkn)
  abund_na <- rowSums(coast_sp_na)
  abund_ph <- rowSums(coast_sp_ph)
  abund_pa <- rowSums(coast_sp_pa)
  abund_mh <- rowSums(coast_sp_mh)
  
  rel_abund <- data.frame(Filterer = abund_filt, Predator = abund_pred, Omnivore = abund_omni,
                          Shredder = abund_shred, Scraper = abund_scrap, Collector = abund_coll, Unknown = abund_un, 
                          NA_FFG = abund_na, `Herbivore Piercer` = abund_ph, Parasite = abund_pa, `Macrophyte Herbivore` = abund_mh,
                          Season = coast_data$Season)
  
  season_rel_abund <- rel_abund %>%
    filter(Season == filter_season) %>%
    select(Filterer:Macrophyte.Herbivore)
  
  return(season_rel_abund)
}

early_ffg_abund <- season_separator("Early")
mid_ffg_abund <- season_separator("Mid")
late_ffg_abund <- season_separator("Late")

summary(early_ffg_abund %>%
  rowSums())

summary(mid_ffg_abund %>%
  rowSums())

summary(late_ffg_abund %>%
  rowSums())
```



# NMDS

## Assemblage Level
```{r}
early_nmds <- metaMDS(sqrt(early_sp), k = 2, autotransform = FALSE)
mid_nmds <- metaMDS(sqrt(mid_sp), k = 2, autotransform = FALSE)
late_nmds <- metaMDS(sqrt(late_sp), k = 2, autotransform = FALSE)
```

## FFG Level
```{r}
early_ffg_nmds <- metaMDS(sqrt(early_ffg_abund), k = 2, autotransform = FALSE)
mid_ffg_nmds <- metaMDS(sqrt(mid_ffg_abund), k = 2, autotransform = FALSE)
late_ffg_nmds <- metaMDS(sqrt(late_ffg_abund), k = 2, autotransform = FALSE)
```

# Vector Fitting

## Vector Fitting Function
```{r}
vector_fn <- function(vector_fit) {
  vector_arrows_df <- data.frame(vector_fit$vectors[1])
  vector_data <- cbind(vector_arrows_df, vector_fit$vectors[2], vector_fit$vectors[3], vector_fit$vectors[4])
  
  vector_data <- vector_data %>%
    mutate(Names = rownames(vector_data)) %>%
    select(Names, everything())
  
  factor_data <- data.frame(vector_fit$factors[2], vector_fit$factors[3], vector_fit$factors[4])
  
  factor_data <- factor_data %>%
    mutate(Names = rownames(factor_data)) %>%
    select(Names, everything())
  
  return_list <- list(vector_data, factor_data)
  
  return(return_list)
}
```

## Assemblage Vector Fitting

```{r}
#Early Vector Fitting

early_vector_fit <- envfit(ord = early_nmds, env = early_env)

early_vector_scores <- vector_fn(early_vector_fit)
early_vectors <- early_vector_scores[[1]]
early_factors <- early_vector_scores[[2]]

#Mid Vector Fitting

mid_vector_fit <- envfit(ord = mid_nmds, env = mid_env)

mid_vector_scores <- vector_fn(mid_vector_fit)
mid_vectors <- mid_vector_scores[[1]]
mid_factors <- mid_vector_scores[[2]]

#Late Vector Fitting
late_vector_fit <- envfit(ord = late_nmds, env = late_env)

late_vector_scores <- vector_fn(late_vector_fit)
late_vectors <- late_vector_scores[[1]]
late_factors <- late_vector_scores[[2]]
```

## FFG Vector Fitting

```{r}
#Early 
early_ffg_vector_fit <- envfit(ord = early_ffg_nmds, env = early_env)

early_ffg_vector_scores <- vector_fn(early_ffg_vector_fit)
early_ffg_vectors <- early_ffg_vector_scores[[1]]
early_ffg_factors <- early_ffg_vector_scores[[2]]

#Mid
mid_ffg_vector_fit <- envfit(ord = mid_ffg_nmds, env = mid_env)

mid_ffg_vector_scores <- vector_fn(mid_ffg_vector_fit)
mid_ffg_vectors <- mid_ffg_vector_scores[[1]]
mid_ffg_factors <- mid_ffg_vector_scores[[2]]

#Late
late_ffg_vector_fit <- envfit(ord = late_ffg_nmds, env = late_env)

late_ffg_vector_scores <- vector_fn(late_ffg_vector_fit)
late_ffg_vectors <- late_ffg_vector_scores[[1]]
late_ffg_factors <- late_ffg_vector_scores[[2]]
```

# Bubble Plots

## Plotting function

```{r}
nmds_plot_fn <- function(nmds_data, vector_data = NA, season_env = NA, title)
{
  
  nmds_points <- data.frame(nmds_data$points)
  
  if (is.na(vector_data) && is.na(season_env))
  {
    p <- ggplot(nmds_points, aes(MDS1, MDS2)) + 
          geom_point(size = 8.5) 
  } else
  {
    nmds_points$PCT_FN <- season_env$pct_fines
    nmds_arrows <- data.frame(scores(vector_data, display = "vector"))
    nmds_arrows$Names <- rownames(nmds_arrows)
    
    p <- ggplot() + 
      geom_point(data = nmds_points, aes(MDS1, MDS2, size = PCT_FN), shape = 1) + 

      scale_size_area(max_size = 10) + 
      geom_segment(data = nmds_arrows, aes(x = 0, xend = NMDS1, y = 0, yend = NMDS2), 
                 arrow = arrow(length = unit(0.25, "cm"))) + 
      geom_text_repel(data = nmds_arrows, aes(x = NMDS1, y = NMDS2, label = Names)) + 
      guides(size = guide_legend(title = "% Fines")) 
  }
  
  p + theme_bw() + 
      theme(panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(), 
          axis.text = element_text(size = 12),
          axis.title = element_text(size = 16),
          plot.title = element_text(size = 16)) +
    ggtitle(paste(title, ", Stress =", round(100 * nmds_data$stress, 1)))
}
```

## NMDS plots (no vectors)

### Assemblage
```{r}
nmds_plot_fn(early_nmds, title = "NMDS on Coastal Streams Assemblage in the Early Season \n(N = 41)")
nmds_plot_fn(mid_nmds, title = "NMDS on Coastal Streams Assemblage in the Mid Season \n(N = 41)")
nmds_plot_fn(late_nmds, title = "NMDS on Coastal Streams Assemblage in the Late Season \n(N = 42)")
```

### FFG
```{r}
nmds_plot_fn(early_ffg_nmds, title = "NMDS on Coastal Streams FFGs in the Early Season \n(N = 41)")
nmds_plot_fn(mid_ffg_nmds, title = "NMDS on Coastal Streams FFGs in the Mid Season \n(N = 41)")
nmds_plot_fn(late_ffg_nmds, title = "NMDS on Coastal Streams FFGs in the Late Season \n(N = 42)")
```


## NMDS plots (with vectors)

## Assemblage
```{r}
nmds_plot_fn(early_nmds, early_vector_fit, early_env, title = "NMDS on Coastal Streams Assemblage in the Early Season \n(N = 41)")
nmds_plot_fn(mid_nmds, mid_vector_fit, mid_env, title = "NMDS on Coastal Streams Assemblage in the Mid Season \n(N = 41)")
nmds_plot_fn(late_nmds, late_vector_fit, late_env, title = "NMDS on Coastal Streams Assemblage in the Late Season \n(N = 42)")
```

## FFG
```{r}
nmds_plot_fn(early_ffg_nmds, early_ffg_vector_fit, early_env, title = "NMDS on Coastal Streams FFGs in the Early Season \n(N = 41)")
nmds_plot_fn(mid_ffg_nmds, mid_ffg_vector_fit, mid_env, title = "NMDS on Coastal Streams FFGs in the Mid Season \n(N = 41)")
nmds_plot_fn(late_ffg_nmds, late_ffg_vector_fit, late_env, title = "NMDS on Coastal Streams FFGs in the Late Season \n(N = 42)")
```
